Renewal and stability in populations structured by remaining years of life

### 2014-08-28 ThanoRepro/ThanoRepro.tex
edits to first half of text

Direct information on the timing of demographic transitions with respect to
death is only available from long-running panel or linked historical data
studies that are few and far between. Some surveys also ask respondents in various ways about
their perceived future lifetime. Perceived remaining lifetime may underly those
demographic transitions governed to some degree by volition, but this is not
treated here. The present inquiry is related to projected
remaining lifetime, using lifetable methods, which is an approximation of
actual remaining lifetime.
The lifetable methods applied here follow those of \citet{brouard1986structure}, \citet{miller2001increasing}, which redistribute age-classified counts over future
death times according to a fixed mortality schedule. This allows us to make
statements about the remaining lifelines within a current stock of population, under certain assumptions.

Temporal perspectives and rescalings of the lifecourse are abundant in
demography already. It is my view that these efforts have enriched the
discipline by producing new insights and questions much more than they have obscured it by
producing conundrums. Likely some demographic phenomena are best
described as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree. Fertility and
reproduction are plausibly a function of both age perspectives, and the question
of how to partition such forces and events over time is a new conundrum that I
do not tackle here. The model presented here should therefore not be interpreted
as a suggestion for a better way to model or project population, but as an alternative way to tell the
story of population renewal within the confines of a philosophical bubble.
Thanatological (counting down towards death) and chronological (counting up
from birth) renewal models, such as the Lotka-Euler renewal model, both
approximate the movement of population stocks over time, but from opposite
vantage points. Specifying a model of thanatological renewal expands the toolbox
available to approach complex temporal population structure. I present a set of
preliminary findings to help understand the implications of temporal structure.

\section*{From chronological to thanatological age}

Before hashing out the thanatological renewal model, I present the
underlying data transformation --- a little formula that
gives the probability of dying at some age $y$ years in the future, given
survival to some earlier age, $a$. For age 0 this probability is simply the $d(a)$ column of the
lifetable, standardized to sum to 1. For higher ages, we simply condition on survival:\footnote{In the reliability literature, this
our concept of $d(y|a)$ is denoted by \begin{equation}
X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current age attained.}

@@

\begin{equation}
d_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{l_a + l_{a+1}} \quad\quad \text{,}
\end{equation}

\noindent where $\omega$ is the highest age.\footnote{In later formulas we just
use $l_a$ in the denominator to reduce clutter, but the above sum over $d_a$
will eliminate rounding error in case one is working from a published life table.} One
produces in this way a probability density function over future death times for
each age, and the $d_{y,a}$- weighted average of $y$ for a given age $a$ returns
the familiar remaining life expectancy column of the lifetable, $e_a$. Since $1
= \sum_{n=0}^{\omega - a}d_{a+y, a}$ for each $a$, we can use this to decompose
age-classified demographic data. For example, we can break down a population
pyramid into remaining years categories, like in Figure~\ref{fig:USdecomp},
which shows this exercise for 2010 US data. The fun starts when we proceed to
regroup cross-classified data by $y$ instead of collapsing back to $a$. Now
instead of remaining lifetime heterogeneity within each age group we have age heterogeneity within each discrete remaining years of life. Figure~\ref{fig:USrecomp} shows this \textit{re}composition for the same 2010 US data.

\begin{figure}
	\caption{2010 US population}
	\begin{center}
	\begin{subfigure}{.45\textwidth}
		\caption{Chronological age-structure decomposed by thanatological age.}
		\label{fig:USdecomp}
		\includegraphics[scale=.6]{Figures/US2010Age.pdf}
	\end{subfigure}
	~
	\begin{subfigure}{.45\textwidth}
		\caption{Thanatological age-structure decomposed by chronological age.}
		\label{fig:USrecomp}
		\includegraphics[scale=.6]{Figures/US2010Thano.pdf}
	\end{subfigure}
	\\
	\small{*Population and mortality data from the HMD.}
	\end{center}
\end{figure}

If $P_a$ is an (chronological) age-classified discrete population count, we can
derive $P_y$, a thanatological age-classified population count, as follows:

\begin{align}
\label{eq:transform}
P_y =& \sum_{a=0}^\omega P_a \frac{d_{a+y}+d_{a+y+1}}{l_a+l_{a+1}}
\intertext{and of course}
\sum P_y =& \sum P_a
\end{align}

We refer to \eqref{eq:transform} as the thanatological age
transformation\footnote{I have on good authority that Ken Wachter
coined the term \textit{thanatological} age.}, and this complements the notion
of chronological age. \citep{brouard1989mouvements}, and later
\citep{vaupel2009life}, showed that in stationary populations the chronological
and thanatological age structures are identical.
It bears repeating that thanatological age is probabilistic until after death, since it refers to the future according to
particular assumptions about mortality. As such, Figure~\ref{fig:USrecomp} is
projective in nature, a forward-looking glance at a population attrition given the age-structure and
lifetable of a particular moment, whereas Figure~\ref{fig:USdecomp} is reflective in nature, since a population's
chronological age-structure is mostly the fruit of past fertility, but also
migration and to a lesser degree attrition. I explore population structure by remaining years of life
for its own sake in a different paper.\footnote{Poster Session 6 (Friday
mid-day))} Note that any count classified by chronological age can be
reclassified in this way, as long as you have a lifetable that plausibly represents the population whose data you wish to reclassify. This means we can derive thanatological fertility rates, $F_y$, by applying \eqref{eq:transform}
to age-classified birth counts, $B_a$, to get $B_y$ and again to
exposure-to-risk, $E_a$ (here we take exposures from all ages), to get $E_y$ and then dividing:

>>>>

Direct information on the timing of demographic transitions with respect to
death is only available from long-running panel or linked historical data that are few and far between. Some surveys also ask respondents in various ways about
their perceived future lifetime, which may underly those
demographic transitions governed to some degree by volition. The present inquiry
draws only from lifetable-projected remaining lifetime, which
is intended to approximate actual remaining lifetime.
The lifetable methods applied here follow those of \citet{brouard1986structure}
or \citet{miller2001increasing}, which redistribute age-classified counts over
future death times according to a fixed mortality schedule. This allows us to make statements about the remaining lifelines within a current stock of population, under certain assumptions.

Temporal perspectives and rescalings of the lifecourse are abundant in
demography already. It is my view that these efforts have enriched the
discipline by producing new insights and questions much more than they have obscured it by
producing conundrums. Likely some demographic phenomena are best
described as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree. Fertility and
reproduction are plausibly a function of both age perspectives, and the question
of how to partition such forces and events over time is a new conundrum that I
do not tackle here. The model presented here should therefore not be interpreted
as a suggestion for a better way to model or project population, but as an
alternative way to tell the story of population renewal.

Thanatological (counting down towards death) and chronological (counting up
from birth) renewal models, such as the Lotka-Euler renewal model, both
approximate the movement of population stocks over time, but from opposite
vantage points. Specifying a model of thanatological renewal expands the toolbox
available to approach complex temporal population structure. I present a set of
preliminary findings to help understand the implications of temporal structure.

\section*{A lifetable approximation of thanatological age}

A lifetable approach to transforming chronologically classified data into
thanatologically classified data is valid when the given lifetable accurately
describes the future mortality of age-classified population. To make
a projective statement about a current stock of population, one does well to
factor in future changes in mortality, and to allow the trajectory of changes
to vary between ages/cohorts. In the case of a stable population model,
mortality is assumed invariant over time, which means that the same survival
pattern underlies each chronological age class, and we therefore draw from a
single lifetable. Under these conditions, data classified by chronological age
can be decomposed and redistributed into thanatological age classes with a
single mortality pattern. The basic operation is to multiply the probability of
surviving from exact chronological age $a$ to exact age $a+y$,
$\frac{l(a+y)}{l(a)}$, and then dying at exact age $a+y$, $\mu(a+y)$. The
index $y$ in this case means $y$ years in the future. For age 0 this probability
is simply the $d(a)$ column of the lifetable, standardized to sum to 1. For higher ages, simply condition on survival:\footnote{In the reliability literature, our concept of $d(y|a)$ is denoted by \begin{equation} X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current chronological age
attained.}

@@

\begin{equation}
d_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \quad\quad \text{,}
\end{equation}

\noindent where $\omega$ is the highest age. One
produces in this way a probability density function over future death times
for each chronological age, $a$, and the $d(y|a)$- weighted average of $y$ for
a given $a$ returns the familiar remaining life expectancy column of the lifetable,
$e(a)$. Since $1 = \int_0^\infty d(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age. For
example, we can break down a population pyramid into thanatological age classes.
Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data, assuming the
2010 period mortality schedule remains constant. When looking from the
chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this \textit{re}composition for the same 2010
US data.

\begin{figure}
	\caption{2010 US population structure}
	\begin{center}
	\begin{subfigure}{.45\textwidth}
		\caption{Chronological age-structure (years lived) decomposed by
		thanatological age (years left).}
		\label{fig:USdecomp}
		\includegraphics[scale=.6]{Figures/US2010Age.pdf}
	\end{subfigure}
	~
	\begin{subfigure}{.45\textwidth}
		\caption{Thanatological age-structure (years left) decomposed by chronological
		age (years lived).}
		\label{fig:USrecomp}
		\includegraphics[scale=.6]{Figures/US2010Thano.pdf}
	\end{subfigure}
	\\
	\small{*Population and mortality data from the HMD.}
	\end{center}
\end{figure}

If $P_a$ is an chronological age-classified discrete population count, we can
derive $P_y$, a thanatological age-classified population count, as follows:

\begin{align}
\label{eq:transform}
P_y =& \sum_{a=0}^\omega P_a \frac{d_{a+y}+d_{a+y+1}}{L_a+L_{a+1}}
\intertext{and of course}
\sum P_y =& \sum P_a
\end{align}

\citet{brouard1989mouvements}, and later
\citet{vaupel2009life}, showed that in stationary populations the chronological
and thanatological age structures are identical.
It bears repeating that lifetable-derived thanatological age structure is
probabilistic until after death, since it refers to the future according to particular assumptions about mortality. As such, Figure~\ref{fig:USrecomp} is projective in nature, a forward-looking glance at a population attrition given the age-structure and
lifetable of a particular moment, whereas Figure~\ref{fig:USdecomp} is
reflective in nature, since a population's chronological age-structure is mostly the fruit of past fertility, but also
migration and to a lesser degree attrition. I explore population structure by remaining years of life
for its own sake in a different paper.\footnote{Poster Session 6 (Friday
mid-day))} Note that any count classified by chronological age can be
reclassified in this way, as long as you have a lifetable that plausibly represents the population whose data you wish to reclassify. This means we can derive thanatological fertility rates, $F_y$, by applying \eqref{eq:transform}
to age-classified birth counts, $B_a$, to get $B_y$ and again to
exposure-to-risk, $E_a$ (here we take exposures from all ages), to get $E_y$ and then dividing:
